Brief Announcement: An Improved Lower Bound for Dynamic Graph Partitioning

### 2020-01-29 paper/paper.tex
Minor rewrite, comments

\mathchardef\mhyphen="2D
\newcommand{\OPT}{\mathit{OPT}}

@@

Additionally, we contribute to the research on the optimal online algorithm by presenting a $O(k\ell)$-competitive algorithm for the special case
where the communication graph (i.e., request sequence) admits a \emph{perfect partition},
i.e., a (multi-way) cut of cost zero that is also an optimal partition for the offline problem.
%and every algorithm must colocate every communicating pair.
Note that our $\Omega(k\ell)$ lower bound generates the input sequence
so that the optimal offline choice is a perfect partition.
\maciek{Also, it is easy to construct the adversary that forces collocation of requested nodes
\mahmoud{this is already the case in the lower bound proof. Do we still need the last sentence here?}
}

@@

Formally, the online \emph{Balanced Re-Partitioning} problem (OBRP) is defined as
follows. There is a set of $n$ nodes,
initially arbitrarily partitioned into $\ell$~clusters,
each of size~$k$.
Nodes in~$C \subset V$ are \emph{collocated} by the partition $P$
if $P$ assigns them to the same cluster.
I.e.,
$\forall u,v \in C: \mathit{cluster} (u,P) = \mathit{cluster} (v,P)$.
The input to the problem is a sequence of communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t,v_t$ exchange a fixed amount of data.
An online algorithm needs to serve the communication request~$(u_t,v_t)$
as soon as it arrives.
Before serving the request,
the online algorithm
can re-partition,
i.e.,
it may move some of the nodes into clusters different than their current assigned clusters.
We assume that
a~communication request between two collocated nodes costs 0,
the cost of a~communication request between two nodes located in different clusters is~1,
and the cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $ALG$,
its cost,
denoted by $ALG(\sigma)$,
is the total cost of communications and
the cost of migrations on the sequence $\sigma$.

@@

In this section we show a lower bound of $\Omega(k \cdot \ell)$
using an input sequence that allows OPT to move to a perfect partition.
Therefore,
it is a lower bound for OBRP with perfect partition,
which directly implies a lower bound for the (general) OBRP.
The former complements an upper bound in $O(k \cdot \ell)$
for the OBRP with perfect partition which we present in Section \ref{sec:upperbound}.

\begin{theorem}
  The competitive ratio of any deterministic algorithm for OBRP is in $\Omega(k\cdot \ell)$.
\end{theorem}

\begin{proof}
We construct an instance of the problem with $\ell$ clusters 
$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
Let $I(C)$ denote the cluster where nodes in a component $C$ are located initially.
Let ALG be any online algorithm for OBRP.
We construct the input sequence of requests for ALG as follows.
First,
we issue $k-2$ (internal) requests so that ALG can form a component of $k-1$
nodes on the cluster $S_1$.
Recall that internal requests does not incur ALG any cost.
If ALG at any point splits a component then
we issue  requests between the parts as many as it is necessary for ALG to collocate the parts,
as otherwise ALG is not competitive.
Let $x_0$  be the only single node left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
Next,
we issue the request between $x_0$ and $y_0$.
Since this is an external request,
ALG joins them into one component and collocates them in some cluster other than $S_1$.
For this,
ALG moves to a new configuration (i.e., partition)
that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.
Next,
we issue a request between $x_1$ and the largest component $C$ s.t.~$I(C) = I(x_1)$.
ALG must collocate $x_1$ and $C$ on some cluster other than $S_1$ and
consequently replaces $x_1$ with some other (single) node $x_2$.
We repeat issuing requests between the single node $x_i$ on $S_1$ and the largest component $C'$ s.t.~$I(C')=I(x_i)$,
 until there are only two single nodes left that  originate from the same cluster.
I.e.,
two single nodes $x^*, y^*,I(x^*) = I(y^*)$ s.t.~for any other pair of single nodes
$x'$ and $y'$,
we have $I(x') \neq I(y')$.
At this point there are at most $\ell+1$ single nodes left,
otherwise there would be more pairs of single nodes that were initially in the same cluster.
Given this sequence of requests,
the optimal strategy is to migrate $\set{x_0,y_0}$ to the cluster $I(x^*)$ by
 swapping $\set{x_0,y_0}$ with $\set{x^*,y^*}$.
Hence,
OPT pays $4$ for node migrations and
ALG pays at least one for each node in the sequence $X := x_0, x_1,\dots$.
We exclude at most $(k-1) + ( \ell+1)$ nodes out of $k.\ell$ nodes,
therefore $|X| \geq k.\ell - k - \ell \in \Omega(k\cdot\ell)$.
\end{proof}

@@

\begin{theorem}	\label{thm:upperbound}
    Algorithm \ref{alg:ppl} reaches the final partition $P_F$ and it is $(2\cdot k\cdot\ell)$-competitive.
\end{theorem}
\begin{proof}
      On each inter-cluster request,
     the algorithm enumerates all $\ell$-way partitions of components
     that are in the same (closest) distance of $P_I$.
     That is, 
     once it reaches $\Delta = \mathit{dist} (P_{t}, P_I)$,
     it does not move to a partition
     $P_{t'}, t'>t,\mathit{dist} (P_{t'}, P_I) > \Delta$
     before it enumerates all partitions at distance $\Delta$.
     Therefore,
     PPL eventually reaches $\Delta=\OPT$ and the partition $P_F$.
%    including the request that completes revealing of all components that are collocated in $P_F$.
    There are at most $(k-1)\cdot\ell < k\cdot\ell $ calls   to $\mathit{re\mhyphen partition}$
     (i.e., the number of internal edges in $P_F$).
    By Lemma \ref{lemma:rebalancecost},
    each costs at most $2\cdot\OPT$.
    The total cost is therefore at most $2\cdot\OPT\cdot k\cdot\ell$ which implies the competitive ratio.
    \mahmoud{This is the cost of moving and the cost of remote comm. is not counted.
    	So it is 4-competitive (?)}
 \end{proof}

@@

Let \emph{GREEDY} be the algorithm that on each inter-cluster request
 moves to a perfect partition closest to the current one.
 That is,
 a partition reachable with the least number of node swaps from the current partition.

\maciek{The definition of the algorithm should be: it moves to an arbitrary perfect partition that is the closest from the current partition.}
\maciek{Therefore, I would name it differently. The easiest would be to call it "greedy" with an explaination what we mean. Alternatives: CPP (closest perfect partition)}

\begin{lemma} \label{lemma:k=3}
  The re-balancing cost of GREEDY for $k=3$ is $O(1)$.
  %\maciek{GREEDY = find the cheapest (closest to the current) feasible partition}
  \label{rebalancing-cost}
\end{lemma}

@@

  We distinguish among three configuration types of algorithm's clusters: $C_1, C_2, C_3$. In $C_1$ we have $3$ singleton components (this is also the initial configuration of any cluster). In $C_2$ we have one component of size $2$ and one component of size $1$. Finally, in $C_3$ we have one component of size $3$.

  Consider a request $(u, v)$ and let $U, V$ be their clusters.
  If $U=V$, then the request does not require a reconfiguration.
  Now, we consider cases upon the type of clusters $U$ and $V$.
  Note that this is impossible that either $U$ or $V$ is $C_3$ as otherwise the resulting component would have the size $4$, and this contradicts the existence of the perfect partition.
  If either $U$ or $V$ is $C_1$, then either cluster can fit the merged component, and the rebalance is local within $U$ and $V$, for the cost of at most $3$ migrations.

  Now, we focus on the case where both $U$ and $V$ are $C_2$. Note that $(u,v)$ cannot both belong to a component of size $2$, as the merged component would have size $4$.
  If either $u$ or $v$ belongs to a component of size $2$ then it suffices to exchange components of size $1$ between $U$ and $V$.
  Finally, if $u$ and $v$ belong to components of size $1$, then we must place them in a cluster different from $U$ and $V$.
  Note that in such case, a $C_1$-type cluster exists, as otherwise the input graph would not have a perfect partition. The cost of re-balancing is then at most $12$ \maciek{can show smaller constant than 12, but it does not matter that much}.
\end{proof}

\begin{theorem}
	Algorithm \emph{GREEDY} is $O(\ell)$-competitive when $k=3$.
\end{theorem}

@@

\begin{theorem} \label{cor:k=3}
	There exists $O(\ell)$-competitive algorithm for the general online problem when $k=3$. 
\end{theorem}
\begin{proof}
	Consider any phase and assume it terminates upon arrival of some request $r$.
	By Lemma  \ref{lemma:k=3},
	\emph{GREEDY} pays $O(\ell)$ during the phase.
	If OPT stays in the same configuration throughout the phase,
	regardless of its partition at the beginning of the phase,        
	it incurs remote communication cost for at least one request $r'$ (possibly $r'=r$) by the end of the phase.
	Else,
	OPT incurs the cost of moving to new partitions.
	In any case,
	OPT pays at least one during the phase and 
	the competitive ratio is $O(\ell)$.
\end{proof}

>>>>

\mathchardef\mhyphen="2D
\newcommand{\OPT}{\mathit{OPT}}
\newcommand{\ALG}{ALG}
\newcommand{\OBRP}{OBRP}
\newcommand{\PPOBRP}{PPOBRP}

@@

Additionally, we contribute to the research on the optimal online algorithm by presenting a $O(k\ell)$-competitive algorithm for the special case
where the communication graph (i.e., request sequence) admits a \emph{perfect partition},
i.e., a (multi-way) cut of cost zero that is also an optimal partition for the offline problem.
%and every algorithm must colocate every communicating pair.
Note that our $\Omega(k\ell)$ lower bound generates the input sequence
so that the optimal offline choice is a perfect partition.
\maciek{Also, it is easy to construct the adversary that forces collocation of requested nodes}
\mahmoud{this is already the case in the lower bound proof. Do we still need the last sentence here?}
\maciek{In general it is OK to emphasize some parts of the paper in the intro. Not sure if this sentence in particular is that important, though.}

@@

\maciek{I suggest to rename the problem to either: (1) Dynamic Balanced Graph Partition or (2) Online Balanced Graph Partition, as in the SIDMA paper}

Formally, the online \emph{Balanced Re-Partitioning} problem (\OBRP) is defined as
follows. We are given a set of $n$ nodes,
initially arbitrarily partitioned into $\ell$~clusters,
each of size~$k$.
We say that a set of nodes $C \subset V$ is \emph{collocated} in a partition $P$
if $P$ assigns them to the same cluster.
\maciek{We can ommit the formal definition: "I.e.,
$\forall u,v \in C: \mathit{cluster} (u,P) = \mathit{cluster} (v,P)$.". However, if we use the notion of cluster later, then TODO: define cluster}
The input consists of a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a fixed amount of data.
An online algorithm needs to serve the communication request~$(u_t,v_t)$
as soon as it arrives.
Before serving the request,
the online algorithm
can re-partition,
i.e.,
it may move some of the nodes into clusters different than their current assigned clusters.
We assume that
a~communication request between two collocated nodes costs 0,
the cost of a~communication request between two nodes located in different clusters is~1,
and the cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $ALG$,
its cost,
denoted by $ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $ALG$ while serving the sequence $\sigma$.
\maciek{Not an expert on commas in english, but here it feels like we have too much. Use some tool to determine (Grammarly?)}

\maciek{Here introduce a restricted model of \PPOBRP, the restricted variant where the input sequence can be perfectly partitioned}

@@

In this section we show a lower bound of $\Omega(k \cdot \ell)$.
The constructed input sequence allows OPT to move to a perfect partition, and hence
the construction also constitues a lower bound for \PPOBRP.
This complements an upper bound of $O(k \cdot \ell)$
for \PPOBRP that we present in Section \ref{sec:upperbound}.

\begin{theorem}
  The competitive ratio of any deterministic algorithm for \OBRP is in $\Omega(k\cdot \ell)$.
\end{theorem}

\begin{proof}
We construct an instance of the problem with $\ell$ clusters 
$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
Let $I(C)$ denote the cluster where nodes in a component $C$ are located initially.
\maciek{So far components were not introduced; the best place to introduce these would be around the definition of \PPOBRP}
Fix any online algorithm \ALG.
We construct the input sequence of requests for \ALG as follows.
First,
we issue $k-2$ (internal) requests so that \ALG can form a component of $k-1$
nodes on the cluster $S_1$.
Recall that internal requests does not incur any cost for \ALG.
If \ALG at any point splits a component\maciek{define splitting}, then
we continue to issue requests between non-collocated nodes of the component until \ALG collocates the nodes of the component.
If \ALG never collocates them then it is not competitive.

Let $x_0$  be the only single node left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
Next,
we issue the request between $x_0$ and $y_0$.
Since this is an external request,
ALG joins them into one component and collocates them in some cluster other than $S_1$.
For this,
ALG moves to a new configuration (i.e., partition)
that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.

Next,
we issue a request between $x_1$ and the largest component $C$ s.t.~$I(C) = I(x_1)$.
ALG must collocate $x_1$ and $C$ on some cluster other than $S_1$ and
consequently replaces $x_1$ with some other (single) node $x_2$.
We repeat issuing requests between the single node $x_i$ on $S_1$ and the largest component $C'$ s.t.~$I(C')=I(x_i)$,
 until there are only two single nodes left that  originate from the same cluster.
I.e.,
two single nodes $x^*, y^*,I(x^*) = I(y^*)$ s.t.~for any other pair of single nodes
$x'$ and $y'$,
we have $I(x') \neq I(y')$.
At this point there are at most $\ell+1$ single nodes left,
otherwise there would be more pairs of single nodes that were initially in the same cluster.

Given this sequence of requests,
the optimal strategy is to migrate $\set{x_0,y_0}$ to the cluster $I(x^*)$ by
 swapping $\set{x_0,y_0}$ with $\set{x^*,y^*}$.
Hence,
OPT pays $4$ for node migrations and
ALG pays at least one for each node in the sequence $X := x_0, x_1,\dots$.
We exclude at most $(k-1) + ( \ell+1)$ nodes out of $k.\ell$ nodes,
therefore $|X| \geq k.\ell - k - \ell \in \Omega(k\cdot\ell)$.
\end{proof}

@@

\begin{theorem}	\label{thm:upperbound}
    Algorithm \ref{alg:ppl} reaches the final partition $P_F$ and it is $(2\cdot k\cdot\ell)$-competitive.
\end{theorem}
\begin{proof}
      On each inter-cluster request,
     the algorithm enumerates all $\ell$-way partitions of components
     that are in the same (closest) distance of $P_I$.
     That is, 
     once it reaches $\Delta = \mathit{dist} (P_{t}, P_I)$,
     it does not move to a partition
     $P_{t'}$, for any $t'>t,\mathit{dist} (P_{t'}, P_I) > \Delta$,
     before it enumerates all partitions at distance $\Delta$.
     Therefore,
     PPL eventually reaches $\Delta=\OPT$ and the partition $P_F$.
%    including the request that completes revealing of all components that are collocated in $P_F$.
    There are at most $(k-1)\cdot\ell < k\cdot\ell $ calls   to $\mathit{re\mhyphen partition}$
     (i.e., the number of internal edges in $P_F$).
    By Lemma \ref{lemma:rebalancecost},
    each re-partition costs at most $2\cdot\OPT$.
    The total cost is therefore at most $2\cdot\OPT\cdot k\cdot\ell$ which implies the competitive ratio.
    \mahmoud{This is the cost of moving and the cost of remote comm. is not counted.
    	So it is 4-competitive (?)}
 \end{proof}

@@

Let \emph{GREEDY} be the algorithm that on each inter-cluster request
 moves to arbitrary perfect partition \maciek{of input} closest (i.e. reachable in minimum number of node swaps) to the current one.

\begin{lemma} \label{lemma:k=3}
  Every re-balancing cost of GREEDY for $k=3$ is $O(1)$.
  %\maciek{GREEDY = find the cheapest (closest to the current) feasible partition}
  \maciek{We need to add to the assumption that the input has a perfect partition}
  \label{rebalancing-cost}
\end{lemma}

@@

  \maciek{We probably should start with something along: Note that the assumption of perfect partition means that there is no component of size 4. Then refer to it (twice) by some short ref.}

  We distinguish among three configuration types of any algorithm's cluster: $C_1, C_2, C_3$, which we define as follows. In $C_1$ the cluster contains $3$ singleton components (this is also the initial configuration of any cluster). In $C_2$ it contains one component of size $2$ and one component of size $1$. Finally, in $C_3$ it contains one component of size $3$.

  Consider a request $(u, v)$ and let $U, V$ be clusters of these nodes.
  If $U=V$, then the request does not require a reconfiguration and GREEDY incurs no cost.
  Otherwise, we consider cases upon the type of clusters $U$ and $V$.
  Note that this is impossible that either $U$ or $V$ is $C_3$ as otherwise we encounter a component of size $4$ and this would contradict the existence of the perfect partition of the input.
  If either $U$ or $V$ is $C_1$, then either cluster can fit the merged component, and the rebalance is local within $U$ and $V$, for the cost of at most $3$ migrations.

  Finally, we consider the case where both $U$ and $V$ are $C_2$. Note that $(u,v)$ cannot both belong to a component of size $2$, as this would mean that an input sequence includes the component of size $4$, a contradiction. 
  If either $u$ or $v$ belongs to a component of size $2$ then it suffices to exchange components of size $1$ between $U$ and $V$.
  Finally, if $u$ and $v$ belong to components of size $1$, then we must place them in a cluster different from $U$ and $V$.
  Note that in such case, a $C_1$-type cluster $W$ exists, as otherwise the input graph would not have a perfect partition. In this case GREEDY exchanges the nodes $u$ and $v$ with any two nodes of $W$, for the total cost of $2\cdot \alpha$.
\end{proof}

\begin{theorem}
  Algorithm \emph{GREEDY} is $O(\ell)$-competitive when $k=3$.
  \maciek{Assumming the perfect partition of input sequence}
\end{theorem}

@@

\begin{theorem} \label{cor:k=3}
	There exists $O(\ell)$-competitive algorithm for the general online problem when $k=3$. 
\end{theorem}
\begin{proof}
  \maciek{TODO: this proof is incomplete!}
	Consider any phase and assume it terminates upon arrival of some request $r$.
	By Lemma  \ref{lemma:k=3},
	\emph{GREEDY} pays $O(\ell)$ during the phase.
	If OPT stays in the same configuration throughout the phase,
	regardless of its partition at the beginning of the phase,        
	it incurs remote communication cost for at least one request $r'$ (possibly $r'=r$) by the end of the phase.
	Else,
	OPT incurs the cost of moving to new partitions.
	In any case,
	OPT pays at least one during the phase and 
	the competitive ratio is $O(\ell)$.
\end{proof}
